### 2017-05-27 paper/conclusions.tex
polished (all)

In this paper, we have proposed an approach to validation of VDM models by discharging proof obligations (POs) with SMT solvers. In our approach, Overture tool is used as the proof obligation generator, and the Python API of Z3 SMT sover is used for encoding and solving formulas of generated POs. Encode strategies including steps of encoding as well as patterns of encoding are demonstrated and two case studies are reported. From our experiences, it is recognized that discharging POs using SMT solvers is efficient, and the counterexample generation of SMT solvers when the discharging fails can provide useful guidance for correcting inconsistency of VDM models. We also discussed other issues such as applicability of encoding strategies considering the automation of our approach. For the next steps of our approach, we are considering more patterns of encoding and how to organize the patterns of encoding for challenging the automation issue.

>>>>

In this paper, we have proposed an approach to validation of VDM models by discharging proof obligations (POs) with SMT solvers. In our approach, Overture tool is used as the proof obligation generator, and the Python API of Z3 SMT sover is used for encoding and solving formulas of generated POs. Encoding strategies including steps of encoding as well as encoding patterns are demonstrated, and two case studies are reported. From our experiences, it is recognized that discharging POs using SMT solvers is efficient, and the counterexample generation of SMT solvers when the discharging fails can provide useful guidance for correcting inconsistency of VDM models. We also discussed other issues such as applicability of encoding strategies considering the automation of our approach. As the future work, we are gathering and organizing more encoding patterns for challenging the automation issue.

### 2017-05-27 paper/introduction.tex
polished (all)

More specifically, in this approach we encode and prove POs of VDM-SL models by solving SMT formulas with Python API of Z3~\cite{MB:08:ZSS}. Z3 is a popular SMT solver widely used in software verification and its Python API provides flexibility in constructing SMT formulas~\cite{BarFT-SMTLIB}. In order to encode POs, built-in VDM types such as sequences and maps must be characterized by theories in Z3. We conducted case studies and the results showed that our approach can efficiently discharge the significant part of POs of a VDM-SL model. We chose the Overture tool as the PO generator in this paper\footnote{POs generated by Overture and VDMTools are slightly different in the numbers and styles of POs}.

>>>>

More specifically, in this approach we chose the Overture tool as the PO generator. For the POs generated by the Overture tool, we encode and prove POs of VDM-SL models by solving SMT formulas with Python API of Z3~\cite{MB:08:ZSS}. Z3 is a popular SMT solver widely used in software verification and its Python API provides flexibility in constructing SMT formulas~\cite{BarFT-SMTLIB}. In order to encode POs, built-in VDM types such as sequences and maps must be characterized by theories in Z3. We conducted case studies and the results showed that our approach can efficiently discharge the significant part of POs of a VDM-SL model.
%  \footnote{POs generated by Overture and VDMTools are slightly different in the numbers and styles of POs}.

### 2017-05-27 paper/obligations.tex
polished (all)

% Generation information of POs
For a VDM model, a proof obligation is a statement that must be proved in order to ensure the consistency of the model. A PO contains a predicate with its context information as shown below. The context information are code segments of the VDM model that relate to the predicate to be proved.

@@

For the sample VDM-SL model shown in Fig.~\ref{fig:module_sample}, Overture generated ten POs. Here selected POs are demonstrated and explained. PO1 is called ``legal sequence application obligation'' that relates to the sequence involved expression (line 9-10) of the VDM-SL model {\tt CMDS}). PO1 says that for all k in the set from 1 to {\tt len s.commands -1}, k should be always within the bound of defined indexes of the sequence {\tt s.commands}. Also, it can be recognized that if we use {\tt len s.commands} instead of {\tt len s.commands -1} the proof will fail. Note that in VDM, a sequence are indexed from 1 to its length. Also note that in PO1, the state {\tt CMDS'S} is mentioned, and we have to include the state invariant as well though it is not explicitly described in PO1.

@@

PO7 is an operation postcondition satisfiable obligation for {\tt push\_cmd} (line 15-18). PO7 says that for every input {\tt a:CMD} and {\tt oldstate}, if the precondition whch takes inputs {\tt a} and {\tt oldstate} is satisfied, the postcondition which takes inputs {\tt a}, {\tt oldstate}, and {\tt newstate} should be satisfied. Note that instead of the specification of precondition, postcondition, and state in {\tt CMDS} module, special names {\tt oldstate}, {\tt newstate}, and {\tt post\_push\_cmd} are used. In VDM-SL, the {\tt pre\_} and {\tt post\_} prefix represent the precondition and postcondition of a function/operator separately. They are evaluable functions if given appropriate inputs, and return a boolean value. In a satisfiable obligation of an operation in which the state change is usually expected, the name of state (in this case {\tt commands}) represents {\tt newstate} while the same name postfixed with ``~\textasciitilde~'' (in this case {\tt commands\textasciitilde}) represents {\tt oldstate}.

>>>>

% Generation information of POs
For a VDM model, a proof obligation (PO) is a statement that must be proved in order to ensure the consistency of the model. A PO contains a predicate with its context information as shown below. The context information are code segments of the VDM model that relate to the predicate to be proved.

@@

For the sample VDM-SL model shown in Fig.~\ref{fig:module_sample}, Overture generated ten POs. Here selected POs are demonstrated and explained. PO1 is called ``legal sequence application obligation'' that relates to the sequence involved expression (line 9-10) of the VDM-SL model {\tt CMDS}). PO1 says that for all k in the set from 1 to {\tt len s.commands -1}, k should be always within the bound of defined indices of the sequence {\tt s.commands}. Also, it can be recognized that if we use {\tt len s.commands} instead of {\tt len s.commands -1} the proof will fail. Note that in VDM, a sequence are indexed from 1 to its length. Also note that in PO1, the state {\tt CMDS'S} is mentioned, and we have to include the state invariant as well though it is not explicitly described in PO1.

@@

PO7 is an operation postcondition satisfiable obligation for {\tt push\_cmd} (line 15-18). PO7 says that for every input {\tt a:CMD} and {\tt oldstate}, if the precondition whch takes inputs {\tt a} and {\tt oldstate} is satisfied, the postcondition which takes inputs {\tt a}, {\tt oldstate}, and {\tt newstate} should be satisfied. Note that instead of the specification of precondition, postcondition, and state in {\tt CMDS} module, special names {\tt oldstate}, {\tt newstate}, and {\tt post\_push\_cmd} are used. In VDM-SL, the {\tt pre\_} and {\tt post\_} prefix represent the precondition and postcondition of a function/operator separately. They are evaluable functions if given appropriate arguments, and return a boolean value. In a satisfiable obligation of an operation in which the state change is usually expected, the name of state (in this case {\tt commands}) represents {\tt newstate} while the same name postfixed with ``~\textasciitilde~'' (in this case {\tt commands\textasciitilde}) represents {\tt oldstate}.

### 2017-05-27 paper/related.tex
polished (all)

% Development history of VDM POG.
Proof is an essential to guarantee the consistency of VDM models~\cite{978-3-540-19813-0}, and the proof obligation generation of VDM is firstly proposed by B. K. Aichernig and P. G. Larsen~\cite{AL:97:POGV}. Later, A. Ribeiro and P. G. Larsen worked on proof obligation generation and discharging related to termination of recursive functions~\cite{Ribeiro2010}.

% Work related to discharging POs using theorem provers. 
There are several researches focusing on translation from VDM to theorem provers and discharging POs of VDM with theorem provers. S. Agerholm~\cite{Agerholm1996} proposed translation and from VDM-SL to PVS. This work proposed a translation from a subset of VDM-SL to PVS for type checking. S. Maharaj and J. Bicarregui~\cite{632849} used Agerholm's translation to assist the verification of VDM-SL models and refinements. Later, S. D. Vermolen et al.~\cite{Verm:2007:master,Vermolen:2010:PCV:1774088.1774608} utilized the parsing mechanism of Overture tools and developed automated translation from a subset of VDM-SL to HOL theorem prover. Vermolen's work also provided some tactics to support automated discharging of POs. L. D. Cuto et al.~\cite{CFP:14:TVCSAP} used the proof obligation generation mechanism for translating CML, a modeling language combing VDM-SL and CSP, to Isabella/HOL for automated proof. Recently, work on improvement of translating VDM-SL to HOL are also proposed~\cite{CT:15:EOCGTIS}.

% Comparison to our approach
The theorem prover based approaches has common issues: (1) the translation from VDM-SL to theorem provers may be erroneous since the translation applies on the whole VDM-SL model; (2) The proof process is usually complicated and requires an expert; (3) if theorem provers fail to prove a PO, there is little information to indicate that the problem is due to incorrect VDM-SL model, incorrect translation, or incorrect tactics. Compare to our approach that discharges VDM-SL proof obligations with SMT solvers, we adopt the efficiency of SMT solvers and our encoding does not require to be applied on the whole VDM-SL model. Furthermore, the functionality of counterexample generation is helpful to determine where the problem is. However, encoding in our approach is manual at this point and still need improvements.

% Work related to validation/verfication of VDM models with model checking.
On the other hand, verification and validation of VDM-SL models with model checking techniques are recently proposed. K. Lausdahl~\cite{kenneth:ifm2013} proposed a semantics-preserving translation that constructs an Alloy model from a subset of VDM-SL. This work aims to support the validation of implicitly specified VDM-SL models with Alloy. H-H. Lin et al.~\cite{DBLP:conf/ftscs/LinOKA15} proposed an approach that combines SPIN model checker and the VDMJ interpreter for verifying explicitly specified VDM-SL models. Lausdahl's work is similar to our approach since Alloy is based on SAT solvers, but SMT solvers in our approach can solve formulas more efficiently. Both the two works require types involved in a VDM-SL model being bounded, while bounds are not necessary for the encoding in our approach. Also, there are issues for Alloy and SPIN to represent the rich and complex VDM-SL types.

>>>>

% Development history of VDM POG.
Proof is essential to guarantee the consistency of VDM models~\cite{978-3-540-19813-0}, and the proof obligation generation of VDM is firstly proposed by B. K. Aichernig and P. G. Larsen~\cite{AL:97:POGV}. Later, A. Ribeiro and P. G. Larsen worked on proof obligation generation and discharging related to termination of recursive functions~\cite{Ribeiro2010}.

% Work related to discharging POs using theorem provers. 
There are several researches focusing on translation from VDM to theorem provers and discharging POs of VDM with theorem provers. S. Agerholm~\cite{Agerholm1996} proposed a translation from a subset of VDM-SL to PVS for type checking. S. Maharaj and J. Bicarregui~\cite{632849} used Agerholm's translation to assist the verification of VDM-SL models and their refinements. Later, S. D. Vermolen et al.~\cite{Verm:2007:master,Vermolen:2010:PCV:1774088.1774608} utilized the parsing mechanism of Overture tools and developed automated translation from a subset of VDM-SL to HOL theorem prover. Vermolen's work also provided some tactics to support automated discharging of POs. L. D. Cuto et al.~\cite{CFP:14:TVCSAP} used the proof obligation generation mechanism of Overture tools for translating CML, a modeling language combing VDM-SL and CSP, to Isabella/HOL for automated proof. Recently, work on improvement of translating VDM-SL to Isabella/HOL are proposed~\cite{CT:15:EOCGTIS}.

% Comparison to our approach
The theorem prover based approaches has common issues: (1) the translation from VDM-SL to theorem provers may be errorneous since the translation applies on the whole VDM-SL model; (2) The proof process is usually complicated and requires an expert even with the help of proof tactics; (3) Theorem provers provide little information when they fail to discharge POs. Compare to our approach that discharges POs of VDM-SL with SMT solvers, we adopt the efficiency of SMT solvers and the encoding does not require to be applied on the whole VDM-SL model. Furthermore, the counterexample generation of SMT solvers is helpful for locating the problem. However, encoding in our approach is manual at this point and still need improvements.

% Work related to validation/verfication of VDM models with model checking.
On the other hand, verification and validation of VDM-SL models with model checking techniques are recently proposed. K. Lausdahl~\cite{kenneth:ifm2013} proposed a semantics-preserving translation that constructs an Alloy model from a subset of VDM-SL. This work aims to support the validation of implicitly specified VDM-SL models with Alloy. H-H. Lin et al.~\cite{DBLP:conf/ftscs/LinOKA15} proposed an approach that combines SPIN model checker and the VDMJ interpreter for verifying explicitly specified VDM-SL models. Lausdahl's work is similar to our approach since Alloy uses SAT solvers, but SMT solvers in our approach can solve wider formulas with more efficiency. Both Lausdahl's and Lin's work require types involved in a VDM-SL model being bounded, and there are issues for Alloy and SPIN to represent the rich and complex VDM-SL types. Compare to them, bounds are not necessary for SMT solvers in our approach and our encoding patterns can applied to complex types thanks for build-in theories in SMT sovers.

### 2017-05-27 paper/strategy.tex
minor fix encode strategy
polished (all)

% Z3 introduction, API introduction
As mentioned in Section~\ref{section:introduction}, our approach encodes and prove VDM POs with the Z3 SMT solver~\cite{MB:08:ZSS}. Z3 is one of the popular SMT solver wildly used in verification of programs and software systems. Z3 has APIs for major programming languages such as C, Java, and Python. The APIs provide a better way of constructing SMT formulas than SMT-LIB~\cite{BarFT-SMTLIB} because one can adopt the flow control characteristics like loop and if-then-else of programming languages to build and solve formulas in a flexible and smart way. In our approach, we chose Z3's Python API (Z3py) as the encoding environment.

@@

\begin{enumerate}
\item
Determine the context information of the PO: The context information is the definitions specified in the VDM model that are involved in the obligation. The context information may include types, functions, and operations. Note that if a type is involved, its type invariant should be included, too.
\item
Encode context information: this step encodes the context information determined by the first step, and can be divided into several minor steps.
\begin{enumerate}
\item
For a VDM type in the context information, both the type itself and its invariant have to be encoded.
\item
For pre- and post-conditions of a function/operation specified in the VDM model, the encoding is needed if {\tt pre\_<name>} or {\tt post\_<name>} are used in the PO. For example, the postcondition satisfiable obligation (PO7) of the {\tt CMDS} module described in Section~\ref{section:proof-obligations} used both precondition and postcondition.
\item
If a PO only considers a part of the specification of a type, function, or an operations, it is only needed to encode the expressions that are directly used in the PO. In other words, if a type, a function, or an operation is not directly used such as variables of the type, evaluation using the function or the operation, there is no need to encode the full specification of the type, the function, or the operation.
\end{enumerate}
\item
Quantifier elimination and obligation negation
\begin{enumerate}
\item
Obligation negation: In model checking with SMT solvers, the assertion to be checked is usually negated, and the result of unsat is expected and means the proof of the assertion succeeds. For the result of sat which means the assertion can be violated, we can get a model as the counterexample that shows the information of how the assertion is violated. In this work, we also take negation on the PO to be discharged. However, the elimination of forall quantifier should be considered together when applying the negation to the PO.
\item
Elimination of forall quantifiers: POs of VDM usually have forall quantifiers. To solve the formulas with forall quantifiers in Z3, though Z3 accepts formulas with forall quantifiers, there are limitations and we may need to eliminate forall quantifiers that may cause problem before encoding. In this work, forall quantifier usually causes difficulties in encoding and solving when complex types are involved. Obligation negation described above may achieve the elimination of troublesome foall quantifiers. However, obligation negation should be applied carefully not to introduce troublesome forall quantifiers from exists quantifiers.
\end{enumerate}
\end{enumerate}

% More details of encoding types, operators, and expressions
VDM has rich types such as natural number, quote, sequence, and mapping, and there are no direct matching for all VDM types to the types in Z3. For example, to encode a natural number, we need to declare an integer and then add a constraint of greater and equal to zero because there is no natural number type in Z3. Also, the enumeration in Z3 can be used to encode simple quote types. For complex types like sequences, maps, or records, we may use uninterpreted functions or arrays to capture the characteristic and operators of the types we want to encode. At this point, there is not yet a complete and systematic way of encoding VDM types to Z3, what we can do is to choose the encoding carefully based on the types and expressions in a VDM model. However, based on our case studies, there are some patterns of encoding that are frequently used for corresponding expression styles in VDM-SL models.

@@

The above Python code imports the z3 module and create a solver object, then declare an integer {\tt i} with a constraint that {\tt i} is greater and equal to zero. In this case, {\tt i} is the encoding an instance of type {\tt nat} in VDM. Note that we cannot define a user-defined type in Z3 but only can declare every instance as Z3 type and add necessary constraints to achieve the encoding of variables of VDM types. This means that if there are several variables declared as VDM type that Z3 does not have, instances of the same number of VDM variables have to be declared in Z3 with constraints for each instance added to the solver object. The strategies discussed above will be demonstrated with details in Section~\ref{section:case-studies} through the report of case studies.

>>>>

% Z3 introduction, API introduction
As mentioned in Section~\ref{section:introduction}, our approach encodes and proves VDM POs with the Z3 SMT solver. Z3 is one of the popular SMT solver wildly used in verification of programs and software systems. Z3 has APIs for major programming languages such as C, Java, and Python. The APIs provide a better way of constructing SMT formulas than SMT-LIB~\cite{BarFT-SMTLIB} because one can adopt the flow control characteristics like loop and if-then-else of programming languages to build and solve formulas in a flexible and smart way. In our approach, we chose Z3's Python API (Z3py) as the encoding environment.

@@

\begin{enumerate}
\item
Determine the context information of the PO: The context information is the definitions specified in the VDM model that are involved in the obligation. The context information may include types, functions, and operations. Note that if a type is involved, its type invariant should be included, too.
\item
Encode context information: this step encodes the context information determined by the first step, and can be divided into several minor steps.
\begin{enumerate}
\item
For a VDM type in the context information, both the type itself and its invariant have to be encoded.
\item
For pre- and post-conditions of a function/operation specified in the VDM model, the encoding is needed if {\tt pre\_<name>} or {\tt post\_<name>} are used in the PO. For example, PO7 of the {\tt CMDS} module described in Section~\ref{section:proof-obligations} mentioned both precondition and postcondition, so both conditions shall be encoded.
\item
If a PO only considers a part of the specification of a type, function, or an operations, it is only needed to encode the expressions that are directly used in the PO. In other words, if a type, a function, or an operation is not directly mentioned, for example, a variable mentioned with {\tt name:type} style, or a function mentioned with {\tt fun\_name(arguments)} style, there is no need to encode the full specification of the type, the function, or the operation.
\end{enumerate}
\item
Obligation negation and quantifier elimination
\begin{enumerate}
\item
Obligation negation: In model checking with SMT solvers, the assertion to be checked is usually negated, and the result of unsatmeans the proof of the assertion succeeds. For the result of sat which means the assertion can be violated, SMT solvers return a model as the counterexample that shows an example of how the assertion is violated. In this work, we also take negation on the PO to be discharged. However, the elimination of forall quantifier should be considered together when applying the negation to the PO.
\item
Elimination of forall quantifiers: POs of VDM usually have forall quantifiers. To solve the formulas with forall quantifiers in Z3, though Z3 accepts formulas with forall quantifiers, there are limitations applying forall quantifiers in Z3 so that by default we have to eliminate forall quantifiers before encoding. In this work, forall quantifier usually causes difficulties in encoding and solving when complex types are involved. Obligation negation described above may achieve the elimination of troublesome foall quantifiers. On the other hand, obligation negation should be applied carefully not to introduce troublesome forall quantifiers from exists quantifiers.
\end{enumerate}
\item
Encode the predicate of obligation and check for satisfiability.
\end{enumerate}

% More details of encoding types, operators, and expressions
VDM has rich types from natural number and quote to sequence and maps. There are no direct matching for all VDM types to the types in Z3. For example, to encode a natural number, we need to declare an integer and then add a constraint of greater and equal to zero because there is no natural number type in Z3. The enumeration type in Z3 can be used to encode simple quote types. For complex types like sequences, maps, or records, we may use uninterpreted functions or arrays as the base type and capture the characteristic and operators of the VDM types by adding corresponding constraints. At this point, there is not yet a complete and systematic way of encoding VDM types in Z3. What we can do is to choose the encoding carefully based on the types and expressions involved in a PO. Based on our case studies, there are patterns of encoding that are frequently used for corresponding expression styles in VDM-SL models.

@@

The above Python code imports the z3 module and create a solver object, then declare an integer {\tt i} with a constraint that {\tt i} is greater and equal to zero. In this case, {\tt i} is the encoding an instance of type {\tt nat} in VDM. Note that we cannot add constraints on types but on instance of types in Z3. This means that if there are several variables of the same VDM type that Z3 does not have, instances of the same number of variables have to be declared in Z3 with constraints for every instance added to the solver object. The steps and strategies of encoding discussed above will be demonstrated with details in Section~\ref{section:case-studies} through the report of case studies.
